Cognate beginnings to bilingual lexical acquisition

### 2023-05-24 manuscript/manuscript.qmd
Add new corrections to discussion and reorganise project structure

---
title: "Cognate beginnings to bilingual lexical growth"
short-title: "Cognate beginnings"
author:
  - name: Gonzalo Garcia-Castro
    corresponding: true
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - id: cbc
        name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        address: Ramon Trias Fargas 25-27
        city: Barcelona
        region: Spain
        postal-code: "08005"
  - name: Daniela S. Avila-Varela
    orcid: 0000-0002-3518-8117
    affiliations: 
      - ref: cbc
      - id: ul
        name: Universidade de Lisboa
        department: Facultade de Letras
        address: Alameda da Universidade
        city: Lisboa
        region: Portugal
        postal-code: "1600-214"
  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    affiliations: 
      - id: uam 
        name: Universidad Autónoma de Madrid
        department: Departamento de Psicología
        address: Iván Pavlov, 6
        city: Madrid
        region: Spain
        postal-code: "28049"
  - name: Núria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    affiliations: 
      - ref: cbc
date: "May 10th, 2023"
code-repo: "Access the code, data, and analysis at <https://github.com/gongcastro/trajectories>"
keywords:
  - lexical acquisition
  - vocabulary
  - bilingualism
  - item response theory
  - bayesian
csl: "../assets/apa7.csl"
bibliography: "../assets/references.bib"
link-citations: true
editor: source

format:
  docx:
    reference-doc: "templates/reference.docx"
    knitr:
      opts_chunk:
        dev: "png"
        dev-args:
          - "cairo-png"
        dpi: 300
        out.width: "80%"
  pdf:
    documentclass: article
    papersize: letter
    fontsize: 12pt
    linestretch: 1.5
    date-format: full
    template-partials:
    - templates/title.tex
    - templates/before-body.tex
    include-in-header: include-in-header.tex
    header-includes: \usepackage{tipa}
    geometry:
    - top=1in
    - bottom=1in
    - left=1in
    - right=1in
    colorlinks: true
    linkcolor: DarkSlateBlue
    urlcolor: DarkSlateBlue
    citecolor: DarkSlateBlue
    link-citations: true
    biblio-title: References
    biblio-config: false
    cite-method: biblatex
    biblatex-chicago: false
    biblio-style: apa
    biblatexoptions: backend=biber
    block-headings: false
    indent: true

@@

library(targets)

tar_config_set(store = here::here('_targets'),
               script = here::here('_targets.R'))

tar_load_globals()

tar_load(c(bvq_data, childes, items, participants, responses))
tar_load(c(model_fit, model_summary, model_epreds, model_convergence))

# process objects
items <- bvq_data$pool  |>  
    mutate(item = str_remove(item, "cat_|spa_")) |> 
    select(item, language, semantic_category, class) |> 
    right_join(items, by = c("item", "language"))

@@

bayesplot_theme_set(my_theme)

@@

Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous research on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. These studies rely on bilingual samples whose languages are relatively distant in typology: English is a Germanic language, while Spanish is a Romance language, and both languages differ substantially at the grammatical and the phonological level. Little is known about vocabulary acquisition in bilinguals learning a pair of typologically closer languages (e.g., when both are Romance languages).

@@

Not all words occur in the linguistic input with identical frequency, so children accumulate experience with some word-forms at faster rates than with other word-forms. This results in frequent words being acquired earlier in age than less frequent words [e.g., @ambridge2015ubiquity; @goodman2008does; @braginsky2019consistency; @swingley2018quantitative; @hansen2017what; @verhagen2022determinants]. For this reason, lexical frequency plays a central role in most models of lexical growth, especially in the case of accumulator models [@hidaka2013computational; @mollica2017how]. Following this account, words with high lexical frequency like 'flower' [`r format(floor(childes[childes$token=="flower",]$freq_million), big.mark = ",")` times per million words in the CHILDES corpus, @macwhinney2000childes] are predicted to be acquired earlier in age than words with low lexical frequency such as 'taxi' (`r format(floor(childes[childes$token=="taxi",]$freq_million), big.mark = ",")` times per million words). We argue that previous accounts on vocabulary growth based on accumulator models are a convenient framework to study bilingual vocabulary growth. Bilinguals share fundamentally similar learning mechanisms with their monolinguals peers. Developmental divergences between both groups can be understood as the result of the the negotiation between such mechanisms and the particular demands of a dual language input [e.g., @werker2005primir; @curtin2011bilingual]. One of such divergences is the fact that bilinguals' frequency of exposure to a word-form is mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time would need more time to accumulate the same experience with 'flower' and 'taxi' than an English monolingual (who listens 100% of the time to English). would need. In the present study, we incorporate this feature of the bilingual linguistic input to an account of the cognateness effect on word acquisition.

We suggest that the cognateness effect proposed by Floccia et al. and Mitchell et al. facilitates word acquisition by reducing the number of learning instances needed by a word to be acquired, and that this effect takes place *only* once the (cognate) translation of such word has been acquired. We hypothesise that both forms in the translation equivalent (e.g., *gat* and *gato* [Catalan and Spanish for 'cat'], for a Catalan-Spanish bilingual) follow their acquisition trajectory as they would for a monolingual child. As the infant receives speech input, both word-forms accumulate learning instances, which occur a a rate proportional to the lexical frequency of each word-form and the amount of exposure to the language the word-form belongs to. When one of the word-forms has accumulated a certain amount of learning instances, it is acquired. After such event takes places, we hypothesise that the number of learning instances left needed for the acquisition of the other word-form (yet to be acquired) is reduced, resulting in an earlier age-of-acquisition that it would be expected if the acquisition of both word-forms was independent. We further suggest that this facilitation is proportional to the amount of phonological similarity (i.e., cognateness) between both word-forms, in such way that the facilitation is larger within similar-sounding translation equivalents, and negligible for form-dissimilar translation equivalents.

Consider the example of the Catalan-Spanish cognate translation equivalent *gat*-*gato* [cat] (whose word-forms share three phonemes), and of the non-cognate translation equivalent *gos*-*perro* [dog] whose word-forms only share one phoneme. When the child has sufficient lexical knowledge of *gat*, and therefore there is some form-referent association in place, future presentations of the similar-sounding translation *gato* in the presence of the same reference should lead to an increased likelihood of acquisition. In the case of a non-cognate translation equivalent, such facilitation would be unlikely. For instance, even if the child has already acquired the word *gos*, hearing its dissimilar Spanish translation *perro* would hardly lead to an increased probability of acquisition. In line with this account, only the later-acquired form of the translation equivalent would benefit from its cognate status. Given that children are more likely to acquire words from languages to which they are more often exposed to [@david2008individual; @cattani2014how; @thordardottir2006bilingual; @hoff2012dual], the acquisition of words in the language of lower exposure should, on average, be more susceptible to the effect of cognateness. For instance, if the previous child is exposed 80% of the time to English, and 20% of the time to Spanish, words in Spanish (especially low-frequency ones) would show a cognateness facilitation effect, with an earlier age-of-acquisition for cognates than for non-cognates.

@fig-hypotheses provides a graphic, theoretical illustration of this account. We assume the case of a Catalan-Spanish bilingual child who is exposed 65% of the time to Catalan, and 35% of the time to Spanish. On the top panels, we illustrate the cumulative sum of learning instances that a child accumulates across ages for the Catalan word-form (/\textipa{gat}/ or /\textipa{gos}/) and its Spanish translation (/\textipa{"pe.ro}/ or /\textipa{"ga.to}/). We assume an arbitrary, theoretical threshold for the learning instances at which the child is considered to have acquired a word (solid grey line). Points and triangles show the age-of-acquisition at which the amount of learning instances reaches the threshold. We show the predicted age-of-acquisition under two hypothesis: (1) a cognateness facilitation effect that lowers the acquisition threshold after one of the word-forms (in this case Catalan /\textipa{gat}/) has been acquired, and (2) no cognateness facilitation, for which the acquisition of the second word-form occurs unconditionally of the acquisition of the first. On the bottom, we show the predicted probability of acquisition for the Catalan and Spanish words under each hypothesis, as it would be expected to observe in a vocabulary questionnaire like the one in the present study. We assume that such probability starts at 0% at birth, reaches 50% (and grows the fastest) at the age-of-acquisition, and reaches 100% at some points in later ages.

@@

To operationalise participants' exposure rate to a word-form, we used a composite measure that captures the lexical frequency of the word-form (an approximation of how often a participant is exposed to it), and child's amount of exposure to the language such word belongs to. This measure was calculated by multiplying the word-form's lexical frequency with participant's degree of exposure to the language the word belongs to. Lexical frequencies were extracted from the CHILDES corpora [@macwhinney2000childes; @sanchez2019childes] (see Appendix A for more details). Participants' degree of exposure were reported by caregivers in a language exposure questionnaire they filled before the vocabulary checklist. Caregivers provided a proportion of the time the child was estimated to listen to any language they were spoken to regularly (described in the methods section). The resulting measure is a exposure-weighted lexical frequency measure, which was included in analyses as a proxy of how often bilingual participants were exposed to each word-form.

*Cognateness* was defined as a continuous measure of phonological similarity between translation equivalents. We quantified the phonological similarity between each pair of Catalan-Spanish word-forms by computing the inverse of the normalised Levenshtein distance between their X-XSAMPA phonetic transcriptions [@levenshtein1966binary; @wells1995computer]. This score is computed by counting the number of insertions, deletions and replacements needed by both transcriptions to become identical, dividing the resulting value by the length of the longest transcription, and finally subtracting this value from one. This results in a proportion that indicates how much the two phonetic transcriptions of the translation equivalent are similar to each other, ranging from zero (no similarity at all) to one (both transcriptions are identical) [see @floccia2018introduction; @fourtassi2020growth; @laing2022phonological for similar approaches]. For instance, this measure of phonological similarity returns 0% for *ocell* /`uceL`/ and *pájaro* /`paxa4o`/, Catalan and Spanish for 'bird', and returns 50% for *lluna* /`Lun5`/ and *luna* /`luna`/, Catalan and Spanish for 'moon' (see Appendix A for more details).

@@

n_items_version <- bvq_data$pool |>
    unnest(version) |> 
    count(version, language) |> 
    pivot_wider(names_from = language,
                values_from = n) |> 
    clean_names() |> 
    mutate(version = tolower(version))

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the formr platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language exposure questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. both lists included items from a diverse sample of `r n_categories` semantic/functional categories. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language, roughly following a one-to-one mapping (e.g., the same participant responded to both *gos* and *perro*, Catalan and Spanish for 'dog') (see @tbl-items for a summary of the questionnaire items).

@@

dates <- participants$time_stamp |> 
    range(na.rm = TRUE) |> 
    as_date()

@@

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between the `r format(dates[1], "%dth of %B, %Y")` and the `r format(dates[2], "%dth of %B, %Y")`: `r n_participants[1]` of those participants participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Recurrent participants provided responses with a minimum of `r as.numeric(diff_responses$diff_min)` days between responses, and a maximum of `r as.numeric(diff_responses$diff_max)`, and were always allocated to the same questionnaire list (A, B, C, or D). Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra), and were contacted by e-mail or phone if their child was aged between `r floor(min(responses$age))` and `r ceiling(max(responses$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan (see @tbl-participants) for a more detailed description of the sample). In total, `r n_trilingual$n[2]` participants (`r percent(n_trilingual$prop[2], accuracy = 0.01)`) participants were reported to be exposed to a third language other than Catalan and Spanish. All families provided informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a token of appreciation.

@@

We used the highest self-reported educational attainment of parents or caregivers as a proxy of participants' socio-economic status (SES). This information was provided by each parent or caregiver by selecting one of six possible alternatives in line with the current educational system in Spain: *sense escolaritzar/sin escolarizar* [no education], *educació primària/educación primaria* [primary school], *educació secundària/educación secundaria* [secondary school], *batxillerat/bachillerato* [complementary studies/high school], *cicles formatius/ciclos formativos* [vocational training], and *educació universitària/educación universitaria* [university degree]. Most families reported university studies (`r participants_edu["University"]`, `r percent(participants_edu["University"]/sum(participants_edu))`), followed by families were the highest educational attainment were vocational studies (`r table(participants$edu_parent)["Vocational"]`, `r percent(participants_edu["Vocational"]/sum(participants_edu))`), complementary studies (`r participants_edu["Complementary"]`,  `r percent(participants_edu["Complementary"]/sum(participants_edu))`), secondary education (`r participants_edu["Secondary"]`, `r percent(participants_edu["Secondary"]/sum(participants_edu))`), no formal education (`r participants_edu["No education"]`, <1%) and primary education (`r participants_edu["Primary"]`, <1%).

@@

rope_interval <- percent(c(lower = -0.1/4, upper = 0.1/4), accuracy = 0.1)

@@

r_version <- glue(r_info$R.version$major, 
                  r_info$R.version$minor, 
                  .sep = ".")
```

@@

```{r results-values}
#| label: results-values
#| echo: false
rhats_df <- model_convergence |> 
    dplyr::filter(str_detect(.variable, "b_|sd_")) |> 
    drop_na()

@@

model_summary |>
    select(.variable_name, .median, .lower, .upper, .rope_overlap, .type) |>
    mutate(across(.median:.upper,
                  \(x) ifelse(.type==intercept_label, plogis(x), x/4))) |> 
    gt(groupname_col = ".type", 
       rowname_col = ".variable_name") |>
    fmt_percent(.rope_overlap, decimals = 3) |> 
    fmt_number(c(.median:.upper), decimals = 3) |>
    cols_merge(c(.lower, .upper),
               pattern = "[{1}, {2}]") |>
    cols_label(
        .variable_name = "Parameter",
        .median = "Median",
        .lower = md("95% HDI"),
        .rope_overlap = md("ROPE prob.")
    ) |>
    tab_style(cell_text(weight = "bold"),
              cells_column_labels(columns = 1:5)) |>
    tab_style(cell_text(align = "left"),
              cells_title(groups = "subtitle")) |> 
    tab_style(cell_text(size = "small"), 
              list(cells_body(),
                   cells_row_groups(),
                   cells_stub())) |> 
    tab_style(cell_text(style = "italic"),
              cells_row_groups())
```

@@

This study investigated the impact of cognateness on the early bilingual lexicon. We used Bayesian Item Response Theory to model the acquisition trajectories of a large sample of Catalan and Spanish words, estimating the effect of cognateness on the probability of acquisition. This model corrected for participants' age, word-form length (number of phonemes), and a novel measure of participants' exposure rate to each word. Exposure rates were calculated as a language exposure-weighted lexical frequency score in which each word-form's lexical frequency was corrected by the degree to which the participant was exposed to each language. Overall, we found that cognates (i.e., phonologically similar translation equivalents) were acquired earlier than non-cognates. This effect was mediated by exposure rate: low-exposure word-forms benefited from their cognate status, whereas high-exposure word-forms did not. Capitalising on accumulator models of language acquisition, we provide a theoretical account of bilingual lexical acquisition in which parallel activation plays a central role during the consolidation of early representations in the bilingual lexicon, and in which the dynamics of co-activation between translation equivalents results in an earlier age-of-acquisition.

>>>>

---
title: "Cognate beginnings to bilingual lexical acquisition"
short-title: "Cognate beginnings"
author:
  - name: Gonzalo Garcia-Castro
    corresponding: true
    orcid: 0000-0002-8553-4209
    email: gonzalo.garciadecastro@upf.edu
    affiliations:
      - id: cbc
        name: Universitat Pompeu Fabra
        department: Center for Brain and Cognition
        address: Ramon Trias Fargas 25-27
        city: Barcelona
        region: Spain
        postal-code: "08005"
  - name: Daniela S. Avila-Varela
    orcid: 0000-0002-3518-8117
    affiliations: 
      - ref: cbc
      - id: ul
        name: Universidade de Lisboa
        department: Facultade de Letras
        address: Alameda da Universidade
        city: Lisboa
        region: Portugal
        postal-code: "1600-214"
  - name: Ignacio Castillejo
    orcid: 0000-0001-7445-0416
    affiliations: 
      - id: uam 
        name: Universidad Autónoma de Madrid
        department: Departamento de Psicología
        address: Iván Pavlov, 6
        city: Madrid
        region: Spain
        postal-code: "28049"
  - name: Núria Sebastian-Galles
    orcid: 0000-0001-6938-2498
    affiliations: 
      - ref: cbc
date: "May 10th, 2023"
code-repo: "Access the code, data, and analysis at <https://github.com/gongcastro/trajectories>"
keywords:
  - lexical acquisition
  - vocabulary
  - bilingualism
  - item response theory
  - bayesian
csl: "../assets/apa7.csl"
bibliography: "../assets/references.bib"
link-citations: true
editor: source

format:
  docx:
    reference-doc: "templates/reference.docx"
    knitr:
      opts_chunk:
        dev: "png"
        dev-args:
          - "cairo-png"
        dpi: 300
        out.width: "80%"
  pdf:
    papersize: letter
    fontsize: 12pt
    linestretch: 1.5
    date-format: full
    template-partials:
    - templates/title.tex
    - templates/before-body.tex
    include-in-header:
    - templates/include-in-header.tex
    header-includes: 
    - \usepackage{tipa}
    - \usepackage{nowidow}
    geometry:
    - top=1in
    - bottom=1in
    - left=1in
    - right=1in
    colorlinks: true
    linkcolor: DodgerBlue
    urlcolor: DodgerBlue
    citecolor: DodgerBlue
    link-citations: true
    biblio-config: false
    cite-method: biblatex
    biblatex-chicago: false
    biblio-style: apa
    biblatexoptions: backend=biber
    block-headings: false
    indent: true

@@

targets::tar_config_set(store = here::here('_targets'),
                        script = here::here('_targets.R'))

targets::tar_load_globals()

targets::tar_load(c(bvq_data, childes, items,
                    participants, responses))

targets::tar_load(c(model_fit, model_summary, 
                    model_epreds, model_convergence))

library(gt)
library(ggplot2)
library(patchwork)

# process objects
items <- bvq_data$pool  |>  
    mutate(item = gsub("cat_|spa_", "", item)) |> 
    select(item, language, semantic_category, class) |> 
    right_join(items, by = c("item", "language"))

@@

@@

Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous research on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilinguala]. These studies rely on bilingual samples whose languages are relatively distant in typology: English is a Germanic language, while Spanish is a Romance language, and both languages differ substantially at the grammatical and the phonological level. Little is known about vocabulary acquisition in bilinguals learning a pair of typologically closer languages (e.g., when both are Romance languages).

@@

Not all words occur in the linguistic input with identical frequency, so children accumulate experience with some word-forms at faster rates than with other word-forms. This results in frequent words being acquired earlier in age than less frequent words [e.g., @ambridge2015ubiquity; @goodman2008doesa; @braginsky2019consistency; @swingley2018quantitative; @hansen2017what; @verhagen2022determinants]. For this reason, lexical frequency plays a central role in most models of lexical growth, especially in the case of accumulator models [@hidaka2013computational; @mollica2017how]. Following this account, words with high lexical frequency like 'flower' [`r format(floor(childes[childes$token=="flower",]$freq_million), big.mark = ",")` times per million words in the CHILDES corpus, @macwhinney2000childes] are predicted to be acquired earlier in age than words with low lexical frequency such as 'taxi' (`r format(floor(childes[childes$token=="taxi",]$freq_million), big.mark = ",")` times per million words).

We argue that previous accounts on vocabulary growth based on accumulator models are a convenient framework to study bilingual vocabulary growth. Bilinguals share fundamentally similar learning mechanisms with their monolinguals peers. Developmental divergences between both groups can be understood as the result of the the negotiation between such mechanisms and the particular demands of a dual language input [e.g., @werker2005primir; @curtin2011bilinguala]. One of such divergences is the fact that bilinguals' frequency of exposure to a word-form is mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time would need more time to accumulate the same experience with 'flower' and 'taxi' than an English monolingual (who listens 100% of the time to English) would need.

In the present study, we incorporate this feature of the bilingual linguistic input to an account of the cognateness effect on word acquisition. We suggest that when a bilingual child is exposed to a given word, they not only activate its corresponding lexical representation, but also that of its translation equivalent. This co-activation is proportional to the form similarity between both word-forms. When this parallel activation is in place, each exposure to one of the word-forms may count as a learning instance for both members of the translation equivalent, in a degree determined by their cognateness. Cognate translation equivalents, for which stronger co-activation is expected, accumulate learning instances faster than non-cognates, for which weaker or no co-activation is expected. Finally,  when the child has accumulator some threshold number of learning instances with any of these word-forms, its lexical representation reaches some degree of consolidation that leads to a non-linear change in the child's observed learning outputs, like the probability of such word being reported as "acquired" by caregivers, or by stronger naming effects in experimental tasks using the Visual World Paradigm [e.g., @swingley2002recognition]. Because , lexical representations of cognate words accumulate learning instances faster than those of non-cognate words, the former would consolidate earlier in age.

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}-\textipa{"ga.to}/ [cat], whose word-forms share three phonemes. When the child is exposed to /\textipa{"gat}/, they might co-activate /\textipa{"ga.to}/ in parallel, due to their phonological similarity. If this exposure to the word form occurs in a convenient referential context (e.g., in the presence of a cat), this might count as a learning instance for both active word-forms. Consider now the case of the non-cognate translation equivalent /\textipa{"gos}-\textipa{"pe.ro}/. Because of their low phonological similarity, the amount of co-activate of both words might be low, which might leads to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for  /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ are predicted to benefit from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ are not. This facilitation effect may not be equivalent for /\textipa{"gat}-\textipa{"ga.to}/. If the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. If parallel activation is in place, the former will activate the latter more often than vice versa. Ultimately, /\textipa{"ga.to}/ should benefit more strongly from its cognate status than /\textipa{"gat}/. For instance, if the previous child is exposed 80% of the time to Catalan, and 20% of the time to Spanish, words in Spanish should, on average, show a stronger cognateness facilitation effect than words in Catalan.

@fig-hypotheses provides a graphic, theoretical illustration of this account. We assume the case of a Catalan-Spanish bilingual child who is exposed 65% of the time to Catalan, and 35% of the time to Spanish. We illustrate the cumulative sum of learning instances that a child accumulates across ages for the Catalan word-form (/\textipa{gat}/ or /\textipa{gos}/) and its Spanish translation (/\textipa{"pe.ro}/ or /\textipa{"ga.to}/). We assume an arbitrary, theoretical threshold for the learning instances at which the child is considered to have acquired a word (solid grey line). Points and triangles show the age-of-acquisition at which the amount of learning instances reaches the threshold. We show the predicted age-of-acquisition under two hypothesis: (1) a cognateness facilitation effect through parallel activation, in which both word-forms co-activate each other accumulating learning instances faster, and (2) no cognateness facilitation, for which both word-forms accumulate learning instances independently.

@@

To operationalise participants' exposure rate to a word-form, we used a composite measure that captures the lexical frequency of the word-form (an approximation of how often a participant is exposed to it), and child's amount of exposure to the language such word belongs to. This measure was calculated by multiplying the word-form's lexical frequency with participant's degree of exposure to the language the word belongs to. Lexical frequencies were extracted from the CHILDES corpora [@macwhinney2000childes; @sanchez2019childesdb] (see Appendix A for more details). Participants' degree of exposure were reported by caregivers in a language exposure questionnaire they filled before the vocabulary checklist. Caregivers provided a proportion of the time the child was estimated to listen to any language they were spoken to regularly (described in the methods section). The resulting measure is a exposure-weighted lexical frequency measure, which was included in analyses as a proxy of how often bilingual participants were exposed to each word-form.

*Cognateness* was defined as a continuous measure of phonological similarity between translation equivalents. We quantified the phonological similarity between each pair of Catalan-Spanish word-forms by computing the inverse of the normalised Levenshtein distance between their X-XSAMPA phonetic transcriptions [@levenshtein1966binary; @wells1995computercoding]. This score is computed by counting the number of insertions, deletions and replacements needed by both transcriptions to become identical, dividing the resulting value by the length of the longest transcription, and finally subtracting this value from one. This results in a proportion that indicates how much the two phonetic transcriptions of the translation equivalent are similar to each other, ranging from zero (no similarity at all) to one (both transcriptions are identical) [see @floccia2018introduction; @fourtassi2020growth; @laing2022phonological for similar approaches]. For instance, this measure of phonological similarity returns 0% for *ocell* /`uceL`/ and *pájaro* /`paxa4o`/, Catalan and Spanish for 'bird', and returns 50% for *lluna* /`Lun5`/ and *luna* /`luna`/, Catalan and Spanish for 'moon' (see Appendix A for more details).

@@

n_items_version <- bvq_data$pool |>
    unnest(version) |> 
    count(version, language) |> 
    pivot_wider(names_from = language,
                values_from = n) |> 
    janitor::clean_names() |> 
    mutate(version = tolower(version))

@@

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire [@garcia-castro2023bvq]. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the formr platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language exposure questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. both lists included items from a diverse sample of `r n_categories` semantic/functional categories. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language, roughly following a one-to-one mapping (e.g., the same participant responded to both *gos* and *perro*, Catalan and Spanish for 'dog') (see @tbl-items for a summary of the questionnaire items).

@@

dates <- participants$time_stamp |> 
    range(na.rm = TRUE) |> 
    lubridate::as_date()

@@

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between the `r format(dates[1], "%dth of %B, %Y")` and the `r format(dates[2], "%dth of %B, %Y")`: `r n_participants[1]` of those participants participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Recurrent participants provided responses with a minimum of `r as.numeric(diff_responses$diff_min)` days between responses, and a maximum of `r as.numeric(diff_responses$diff_max)`, and were always allocated to the same questionnaire list (A, B, C, or D). Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra), and were contacted by e-mail or phone if their child was aged between `r floor(min(responses$age))` and `r ceiling(max(responses$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan (see @tbl-participants) for a more detailed description of the sample). In total, `r n_trilingual$n[2]` participants (`r scales::percent(n_trilingual$prop[2], accuracy = 0.01)`) participants were reported to be exposed to a third language other than Catalan and Spanish. All families provided informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a token of appreciation.

@@

We used the highest self-reported educational attainment of parents or caregivers as a proxy of participants' socio-economic status (SES). This information was provided by each parent or caregiver by selecting one of six possible alternatives in line with the current educational system in Spain: *sense escolaritzar/sin escolarizar* [no education], *educació primària/educación primaria* [primary school], *educació secundària/educación secundaria* [secondary school], *batxillerat/bachillerato* [complementary studies/high school], *cicles formatius/ciclos formativos* [vocational training], and *educació universitària/educación universitaria* [university degree]. Most families reported university studies (`r participants_edu["University"]`, `r scales::percent(participants_edu["University"]/sum(participants_edu))`), followed by families were the highest educational attainment were vocational studies (`r table(participants$edu_parent)["Vocational"]`, `r scales::percent(participants_edu["Vocational"]/sum(participants_edu))`), complementary studies (`r participants_edu["Complementary"]`,  `r scales::percent(participants_edu["Complementary"]/sum(participants_edu))`), secondary education (`r participants_edu["Secondary"]`, `r scales::percent(participants_edu["Secondary"]/sum(participants_edu))`), no formal education (`r participants_edu["No education"]`, <1%) and primary education (`r participants_edu["Primary"]`, <1%).

@@

rope_interval <- scales::percent(c(lower = -0.1/4,
                                   upper = 0.1/4),
                                 accuracy = 0.1)

@@

r_version <- glue::glue(r_info$R.version$major, 
                        r_info$R.version$minor, 
                        .sep = ".")
```

@@

```{r results-values}
#| label: results-values
#| echo: false
rhats_df <- model_convergence |> 
    dplyr::filter(grepl("b_|sd_", .variable)) |> 
    drop_na()

@@

model_summary |>
    select(.variable_name, .median, .lower, .upper, .rope_overlap, .type) |>
    mutate(across(.median:.upper,
                  \(x) ifelse(.type==intercept_label, plogis(x), x/4))) |> 
    gt(groupname_col = ".type", 
       rowname_col = ".variable_name") |>
    fmt_percent(.rope_overlap, decimals = 3) |> 
    fmt_number(c(.median:.upper), decimals = 3) |>
    cols_merge(c(.lower, .upper),
               pattern = "[{1}, {2}]") |>
    cols_label(
        .variable_name = "Parameter",
        .median = "Median",
        .lower = md("95% HDI"),
        .rope_overlap = md("*p*(H0)")
    ) |>
    tab_style(cell_text(weight = "bold"),
              cells_column_labels(columns = 1:5)) |>
    tab_style(cell_text(align = "left"),
              cells_title(groups = "subtitle")) |> 
    tab_style(cell_text(size = "small"), 
              list(cells_body(),
                   cells_row_groups(),
                   cells_stub())) |> 
    tab_style(cell_text(style = "italic"),
              cells_row_groups())
```

@@

This study investigated the impact of cognateness on the early bilingual lexicon. We used Bayesian Item Response Theory to model the acquisition trajectories of a large sample of Catalan and Spanish words, estimating the effect of cognateness on the probability of acquisition. This model corrected for participants' age, word-form length (number of phonemes), and a novel measure of participants' exposure rate to each word. Exposure rates were calculated as a language exposure-weighted lexical frequency score in which each word-form's lexical frequency was corrected by the degree to which the participant was exposed to each language. Overall, we found that cognates (i.e., phonologically similar translation equivalents) were acquired earlier than non-cognates. This effect was mediated by exposure rate: low-exposure word-forms benefited from their cognate status, whereas high-exposure word-forms did not. Capitalising on accumulator models of language acquisition, we provided a theoretical account of bilingual lexical acquisition. In this account, parallel activation of the two languages plays a central role during the consolidation of early representations in the bilingual lexicon, and in which the dynamics of co-activation between translation equivalents results in an earlier age-of-acquisition.

This investigation is particularly relevant in the light of two previous findings. First, @floccia2018introduction reported that bilingual toddlers learning two typologically close languages (e.g. shared many cognates, like., English-German) showed larger vocabulary sizes than those learning typologically distant languages (e.g. shared fewer cognates, like English-Mandarin). Second, @mitchell2022cognates found an earlier age-of-acquisition for cognates, compared to non-cognates. The outcomes of both studies pointed to cognateness facilitating word acquisition through parallel activation. But the underpinnings of such effect were unclear: while parallel activation has been extensively described in experimental studies, current paradigms of bilingual word acquisition and word learning are, to a large extent, dissociated from the mechanisms proposed by previous work on word processing. Accumulator models of language acquisition provide a convenient theoretical framework to narrow this gap.
